---
title: "Quarto - instructor notes"
subtitle: "Adapted from: https://cforgaci.github.io/r-socialsci/08-quarto.html"
format: html
editor: 
  markdown: 
    wrap: 72
---

## Create a Quarto file

Quarto comes pre-installed with RStudio (as of v2022.07), so no action
is necessary.

-   `RStudio, click File -> New File -> Quarto Document`

-   Click on `Create Empty Document`

## Create a Quarto file

To control the output, a YAML (YAML Ain’t Markup Language) header is
needed:

```         
---
title: "My Awesome Report"
author: "Bilbo Baggins"
date: ""
format: html
---
```

-   The header is defined by the three hyphens at the beginning (`---`)
    and the three hyphens at the end (`---`).

-   You can add more information about your document in the YAML header,
    that will provide more information about the document or change the
    behaviour of the file.

-   You can also adapt the `format` of the file, to `pdf` or `docx`.

## Markdown syntax

-   Markdown is a popular markup language that allows you to add
    formatting elements to text.

-   editing not visible in markdown -\> markdown converted to other file
    formats.

### Headings

A `#` in front of text indicates to Markdown that this text is a
heading. Adding more `#`s goes makes heading smaller:

```         
# Title
## Section
### Sub-section
#### Sub-sub section
##### Sub-sub-sub section
###### Sub-sub-sub-sub section
```

-   only use a level if the one above is also in use

-   probably not a great practice to have more than 3 subheadings.

-   Already have a title, so start from `##` to create introduction

```         
## Introduction
```

### Formatting

-   `_` or `*` for *italics*

-   `__` or `**` for **bold**

-   \`\``for`code-type\`

-   You can use combination of those things (but do you realy want to?)

```         
## Introduction

This report uses the `tidyverse` **package** along with the *diamonds* dataset, 
which has columns that include:
```

### Lists

```         
## Introduction

This report uses the **tidyverse** package along with the *SAFI* dataset, 
which has columns that include:

- carat
- cut
- price
- color
```

-   You can also create an ordered list using numbers:

```         
1. carat
2. cut
3. price
4. color
```

-   And nested items by tab-indenting:

```         
- carat
  + weight of the diamond
- cut
  + quality of the cut
- price
  + price in US dollars
- color
  + diamond colour, from D (best) to J (worst)
```

Now we can render the document into HTML by clicking the **Render**
button in the top of the Source pane (top left), or use the keyboard
shortcut <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>K</kbd> on Windows and
Linux, and <kbd>Cmd</kbd>+<kbd>Shift</kbd>+<kbd>K</kbd> on Mac.

If you haven't saved the document yet, you will be prompted to do so
when you **Render** for the first time.

## Writing a Quarto report

Now we will add some R code with some data wrangling and visualisation.

We need to make sure packages you need are loaded. We will use
**tidyverse** .

It is not enough to load **tidyverse** from the console, we will need to
load it within our Quarto document.

The same applies to our data.

To load these, we will need to create a **'code chunk'** at the top of
our document (below the YAML header).

A code chunk can be inserted by clicking `Code > Insert Chunk`, or by
using the keyboard shortcuts <kbd>Ctrl</kbd>+<kbd>Alt</kbd>+<kbd>I</kbd>
on Windows and Linux, and <kbd>Cmd</kbd>+<kbd>Option</kbd>+<kbd>I</kbd>
on Mac.

The syntax of a code chunk is:

```{r}
#| echo: fenced
#| label: chunk-name

# Here is where you place the R code that you want to run.
```

An Quarto document knows that this text is not part of the report from
the ```` ``` ```` that begins and ends the chunk. It also knows that the
code inside of the chunk is R code from the `r` inside of the curly
braces (`{}`).

Below the curly braces, you can add code chunk options after the `#|`
sign. In this way, you can for example add a label for the code chunk.

::: callout
Naming a chunk is optional, but recommended. **Each chunk label must be
unique**, and only contain alphanumeric characters and `-`.
:::

```{r}
#| echo: false
#| label: package-data
#| message: false
#| purl: false

library(tidyverse)
library(here)

diamonds_df <- diamonds |>
  select(carat, cut, price, color)
  
```

To load **tidyverse** we will insert a chunk and call it 'setup'.

Since we don't want this code or the output to show in our knitted HTML
document, we add an `#| include: false` option after the curly braces.

```{{r}}
#|label: setup
#|include: false

library(tidyverse)
library(here)

diamonds_df <- diamonds |>
  select(carat, cut, price, color)

```

::: callout
## Important Note!

The file paths you give in a .qmd document, e.g. to load a .csv file,
are relative to the .qmd document, **not** the project root.

As suggested in the Starting with Data episode, we highly recommend the
use of the `here()` function to keep the file paths consistent within
your project.
:::

## Insert table

We will now create a table with some summary statistics about the
dataset.

::: callout-note
It isn't necessary to **Render** your document every time you want to
see the output. Instead you can run the code chunk with the green
triangle in the top right corner of the the chunk, or with the keyboard
shortcuts: <kbd>Ctrl</kbd>+<kbd>Alt</kbd>+<kbd>C</kbd> on Windows and
Linux, or <kbd>Cmd</kbd>+<kbd>Option</kbd>+<kbd>C</kbd> on Mac.
:::

To make sure the table is formatted nicely in our output document, we
will need to use the `kable()` function from the **knitr** package. The
`kable()` function takes the output of your R code and renders it into a
nice looking HTML table. You can also specify different aspects of the
table, e.g. the column names, a caption, etc.

Run the code chunk to make sure you get the desired output.

```{r}
#| purl: false
diamonds_df |>
    filter(!cut %in% c("Fair", "Good")) |>
    group_by(cut, color) |>
    summarize(mean_price = mean(price)) |>
  knitr::kable(col.names = c("Cut", "Color", 
                             "Average Price"))
```

When you are generating a table in Quarto the label should be prefixed
with `tbl-`, e.g. `tbl-diamonds`.

You can add a caption to the chunk options with
`tbl-cap: "Your caption here"`.

```{{r}}
#| label: tbl-diamonds
#| tbl-cap: "A useful description about the table."
#| eval: true
#| echo: true

diamonds_df |>
    filter(!cut %in% c("Fair", "Good")) |>
    group_by(cut, color) |>
    summarize(mean_price = mean(price)) |>
  knitr::kable(col.names = c("Cut", "Color", 
                             "Average Price"))
```

```{r}
#| label: tbl-interview-out
#| echo: false
diamonds_df |>
    filter(!cut %in% c("Fair", "Good")) |>
    group_by(cut, color) |>
    summarize(mean_price = mean(price)) |>
  knitr::kable(col.names = c("Cut", "Color", 
                             "Average Price"))
```

## Customising chunk output

We mentioned using `include: false` in a code chunk to prevent the code
and output from printing in the rendered document. There are additional
options available to customise how the code-chunks are presented in the
output document. The options are entered in the code chunk using the
'hash pipe', `#|`.

| Option | Options | Output |
|----------|----------|-----------------------------------------------------|
| `eval` | `TRUE` or `FALSE` | Whether or not the code within the code chunk should be run. |
| `echo` | `TRUE` or `FALSE` | Choose if you want to show your code chunk in the output document. `echo = TRUE` will show the code chunk. |
| `include` | `TRUE` or `FALSE` | Choose if the output of a code chunk should be included in the document. `FALSE` means that your code will run, but will not show up in the document. |
| `warning` | `TRUE` or `FALSE` | Whether or not you want your output document to display potential warning messages produced by your code. |
| `message` | `TRUE` or `FALSE` | Whether or not you want your output document to display potential messages produced by your code. |
| `fig-align` | `default`, `left`, `right`, `center` | Where the figure from your R code chunk should be output on the page |

::: callout
## Tip

-   The default settings for the above chunk options are all `true`.
-   The default settings can be modified per chunk, or with
    `knitr::opts_chunk$set()`,
-   Entering `knitr::opts_chunk$set(echo = FALSE)` will change the
    default of value of `echo` to `FALSE` for **every** code chunk in
    the document.

The defaults can also be changed in the YAML header with:

```         
---
knitr:
  opts_chunk:
    echo: false
---
```
:::

:::: challenge
## Exercise

Play around with the different options in the chunk with the code for
the table, and re-**Render** to see what each option does to the output.

What happens if you use `eval: false` and `echo: false`? What is the
difference between this and `include: false`?

::: solution
## Solution to Exercise

Create a chunk with `eval: false, echo: false`, then create another
chunk with `include: false` to compare. `eval: false` and `echo: false`
will neither run the code in the chunk, nor show the code in the
rendered document. The code chunk essentially doesn't exist in the
rendered document as it was never run. Whereas `include: false` will run
the code and store the output for later use.
:::
::::

## In-line R code

To use in-line R-code, we use the same backticks that we used in the
Markdown section, with an `r` to specify that we are generating R-code.

The difference between in-line code and a code chunk is the number of
backticks.

For example, today's date is ``` ``r "r Sys.Date()"`` ```, will be
rendered as: today's date is `r Sys.Date()`.

The code will display today's date in the output document (well,
technically the date the document was last rendered).

The best way to use in-line R code, is to minimise the amount of code
you need to produce the in-line output by preparing the output in code
chunks.

Let's say we're interested in presenting the average household size in a
village.

```{r}
#| label: in-line-code
#| purl: false
# create a summary data frame with the mean price by cut
mean_price <- diamonds_df %>%
    group_by(cut) %>%
    summarize(mean_price = mean(price))

# and select the village we want to use
mean_ideal <- mean_price %>%
  filter(cut == "Ideal")
```

Now we can make an informative statement on the means of each cut, and
include the mean values as in-line R-code. For example:

The average price of ideally cut diamonds is
``` ``r "r round(mean_ideal$mean_price, 2)"`` ```\$.

becomes...

The average price of ideally cut diamonds is
`r round(mean_ideal$mean_price, 2)`\$.

Because we are using in-line R code instead of the actual values, we
have created a dynamic document that will automatically update if we
make changes to the dataset and/or code chunks.

## Plots

Plots created in Quarto should have a label prefixed with `fig-`, e.g.
`#| label: fig-fancy-plot`.

```{r}
#| label: fig-diamonds-plot
#| purl: false

diamonds_df %>%
  ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = color), alpha = 0.6)
```

We can also create a caption with the chunk option
`fig-cap: "Caption here"`, and add some nicer labels using the `labs()`
function.

```{{r}}
#| label: fig-diamonds-plot
#| fig-cap: "Yet another diamonds plot"
#| echo: true
#| eval: false

diamonds_df %>%
  ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = color), alpha = 0.6)+ 
  labs(x = "Diamond weight (carat)", y = "Diamond price ($)", color = "Color") 
```

why put prefix in the labels of your tables in figures?

It allows you to cross-reference them in the text of your document, and
it requires that the label is unique, and that they have the correct
prefix.

For example, we can talk about the table we made earlier and reference
it using the label `@tbl-diamonds`, which, when rendered, becomes Table
1.

We can do the same with our figures. For example, `@fig-diamonds-plot`
becomes Figure 1. The number will of course depend on whether any plots
or figures comes before it, but since you just need to reference the
label, there's no need to know what number a specific plot has in a
document.

## Other output options

You can convert Quarto to a PDF or a Word document (among others). Put
`pdf` or `word` in the initial header of the file to indicate the
desired output format.

```         
---
format: word
---
```

::: callout
## Note: Creating PDF documents

Creating .pdf documents may require installation of some extra software.
The R package `tinytex` provides some tools to help make this process
easier for R users. With `tinytex` installed, run
`tinytex::install_tinytex()` to install the required software (you'll
only need to do this once) and then when you render to pdf `tinytex`
will automatically detect and install any additional LaTeX packages that
are needed to produce the pdf document. Visit the [tinytex
website](https://yihui.org/tinytex/) for more information.
:::

::: callout
## Note: Inserting citations into a Quarto file

It is possible to insert citations into a Quarto file using the editor
toolbar. The editor toolbar includes commonly seen formatting buttons
generally seen in text editors (e.g., bold and italic buttons). The
toolbar is accessible by using the settings dropdown menu (next to the
'Render' dropdown menu) to select 'Use Visual Editor', also accessible
through the shortcut 'Crtl+Shift+F4'. From here, clicking 'Insert'
allows 'Citation' to be selected (shortcut: 'Crtl+Shift+F8'). For
example, searching '10.1007/978-3-319-24277-4' in 'From DOI' and
inserting will provide the citation for `ggplot2` [@wickham2016]. This
will also save the citation(s) in 'references.bib' in the current
working directory.

Visit the [R Studio
website](https://rstudio.github.io/visual-markdown-editing/) for more
information. Tip: obtaining citation information from relevant packages
can be done by using `citation("package")`.
:::

## Resources

-   [Markdown tutorial](https://commonmark.org/help/tutorial/)
-   [Official Quarto website](https://quarto.org/) (comprehensive
    resource of tutorials and documentation)
-   [Welcome to
    Quarto](https://www.youtube.com/watch?v=yvi5uXQMvu4&ab_channel=PositPBC) -
    workshop by Posit (former RStudio)
-   [R Markdown: The Definitive
    Guide](https://bookdown.org/yihui/rmarkdown/) - book by the RStudio
    team on R Markdown, the predecessor of Quarto
